---
title: "Diabetes Risk Predictions"
author: "Seif Kungulio"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)


```
<hr style="height:5px; border:none; color:#FF0000; background-color:#FF0000;" />

## Business Understanding
<hr style="height:2px; border:none; color:#006000; background-color:#006000;" />

### Data Collection
The primary dataset used is the **Diabetes Health Indicators Dataset** sourced from **Kaggle**, based on the **2015 Behavioral Risk Factor Surveillance System (BRFSS)**. This dataset includes over **253,000 observations** and **22 attributes**, offering a rich foundation for predictive modeling using demographic, behavioral, and clinical health indicators.

### Data Description
The dataset contains 253,680 records and 22 variables relevant to diabetes risk prediction. The target variable is binary (Diabetes_binary), while the predictors include behavioral, clinical, and demographic factors such as blood pressure, cholesterol, BMI, physical activity, general health, and income. Most variables (17) are binary categorical, with the remaining 5 being numeric or ordinal. This structure supports classification modeling and offers strong coverage of key health indicators needed to identify individuals at risk for diabetes.

### Load the data
```{r}
diabetes.df <- read.csv("diabetes_health_indicators_BRFSS2015.csv")
```
### Check for dimension and structure
```{r}
dim(diabetes.df)
str(diabetes.df)
```
### Check for statistical summary
```{r}
summary(diabetes.df)
```

### Business Objective
The primary business objective of this project is to support the **Texas Department of State Health Services** in enhancing diabetes prevention efforts across the state. By leveraging self-reported health behavior and demographic data, the goal is to develop a predictive model by **2026** that identifies **Texas adults with a 30% or higher risk** of developing diabetes within the next five years. This model aims to facilitate **early detection**, reduce **long-term healthcare costs**, promote **health equity**, and **optimize resource allocation** in vulnerable communities.

### Problem Statement
How can public health agencies in Texas use self-reported behavioral and demographic data to **identify adults at 30%+ risk of developing diabetes within 5 years**, using markers like **HbA1c (>5.7%)** and **BMI (>30)**, and implement targeted interventions by **2026**?.

### Business Success Criteria
The project will be considered successful if it:

- Produces a validated predictive model with practical thresholds for decision-making
- Demonstrates improved identification of high-risk populations compared to current screening protocols
- Supports targeted outreach that aligns with state health equity goals
- Provides actionable insights for stakeholders through dashboards or visual summaries

<br>


## Data Understanding
<hr style="height:2px; border:none; color:#006000; background-color:#006000;" />

## Data Preparation
<hr style="height:2px; border:none; color:#006000; background-color:#006000;" />

## Business Understanding
<hr style="height:2px; border:none; color:#006000; background-color:#006000;" />